---
title: "Strom-data-analysis"
author: "Yash Gupta"
date: "15 April 2020"
output: html_document
---
# Impact of Severe Weather Events on Population Health and Property Damage

1: Synopsis
-----------

The goal of the assignment is to explore the NOAA Storm Database and explore the effects of severe weather events on both population and economy.The database covers the time period between 1950 and November 2011.

The following analysis investigates which types of severe weather events are most harmful on:

1.  Health (injuries and fatalities)
2.  Property and crops (economic consequences)

Information on the Data: [Documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf)

2: Data Processing
------------------

### 2.1: Data Loading

Download the raw data file and extract the data into a dataframe.Then convert to a data.table

Load File. Note: File was downloaded from site [Storm data](https://www.coursera.org/learn/reproducible-research/peer/OMZ37/course-project-2) as a csv.bz2. 
File was unzipped and extracted to working directory

```{r}
library("data.table")
library("ggplot2")
#download the data 
fileUrl<-"https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
if(!file.exists("data"))
        dir.create("data")
#download.file(fileUrl, destfile = "data/stormData.csv.bz2")

#read the file
stormData<-read.csv(bzfile("data/stormData.csv.bz2"))
stormData<-as.data.table(stormData)
```

### 2.2: Examining Data
View Headers

```{r}
#View header
head(stormData)
```

### 2.3: Data Subsetting

Subset the dataset on the parameters of interest. Basically, we remove the columns we don't need for clarity.

``` {r}
# Only use data where fatalities or injuries occurred and keep only reqired column.  
stormData <- stormData[(EVTYPE != "?" & 
             (INJURIES > 0 | FATALITIES > 0 | PROPDMG > 0 | CROPDMG > 0)), c("EVTYPE"
                                                                            , "FATALITIES"
                                                                            , "INJURIES"
                                                                            , "PROPDMG"
                                                                            , "PROPDMGEXP"
                                                                            , "CROPDMG"
                                                                            , "CROPDMGEXP") ]
head(stormData)
```

### 2.4: Converting Exponent Columns into Actual Exponents instead of (-,+, H, K, etc)

Making the PROPDMGEXP and CROPDMGEXP columns cleaner so they can be used to calculate property and crop cost.

```{r}
# Change all damage exponents to uppercase.
cols <- c("PROPDMGEXP", "CROPDMGEXP")
stormData[,  (cols) := c(lapply(.SD, toupper)), .SDcols = cols]

# Map property damage alphanumeric exponents to numeric values.
propDmgKey <-  c("\"\"" = 10^0,
                 "-" = 10^0, 
                 "+" = 10^0,
                 "0" = 10^0,
                 "1" = 10^1,
                 "2" = 10^2,
                 "3" = 10^3,
                 "4" = 10^4,
                 "5" = 10^5,
                 "6" = 10^6,
                 "7" = 10^7,
                 "8" = 10^8,
                 "9" = 10^9,
                 "H" = 10^2,
                 "K" = 10^3,
                 "M" = 10^6,
                 "B" = 10^9)

# Map crop damage alphanumeric exponents to numeric values
cropDmgKey <-  c("\"\"" = 10^0,
                "?" = 10^0, 
                "0" = 10^0,
                "K" = 10^3,
                "M" = 10^6,
                "B" = 10^9)

stormData[, PROPDMGEXP := propDmgKey[as.character(stormData[,PROPDMGEXP])]]
stormData[is.na(PROPDMGEXP), PROPDMGEXP := 10^0 ]

stormData[, CROPDMGEXP := cropDmgKey[as.character(stormData[,CROPDMGEXP])] ]
stormData[is.na(CROPDMGEXP), CROPDMGEXP := 10^0 ]

```

### 2.5: Making Economic Cost Columns

``` {r}
stormData <- stormData[, .(EVTYPE, FATALITIES, INJURIES, PROPDMG, PROPDMGEXP, propCost = PROPDMG * PROPDMGEXP, CROPDMG, CROPDMGEXP, cropCost = CROPDMG * CROPDMGEXP)]
```

### 2.6: Calcuating Total Property and Crop Cost

``` {r}
totalCostDT <- stormData[, .(propCost = sum(propCost), cropCost = sum(cropCost), Total_Cost = sum(propCost) + sum(cropCost)), by = .(EVTYPE)]

totalCostDT <- totalCostDT[order(-Total_Cost), ]

totalCostDT <- totalCostDT[1:10, ]

head(totalCostDT, 5)
```

### 2.7: Calcuating Total Fatalities and Injuries

```{r}
totalInjuriesDT <- stormData[, .(FATALITIES = sum(FATALITIES), INJURIES = sum(INJURIES), totals = sum(FATALITIES) + sum(INJURIES)), by = .(EVTYPE)]

totalInjuriesDT <- totalInjuriesDT[order(-FATALITIES), ]

totalInjuriesDT <- totalInjuriesDT[1:10, ]

head(totalInjuriesDT, 5)
```

3: Results
----------

### 3.1: Events that are Most Harmful to Population Health

Melting data.table so that it is easier to put in bar graph format

```{r}
fatalInjury <- melt(totalInjuriesDT, id.vars="EVTYPE", variable.name = "fatalInjury")
head(fatalInjury, 5)

#create graph
ggplot(fatalInjury, aes(fill = fatalInjury, x = reorder(EVTYPE,-value), y = value))+
        geom_bar(position = "dodge", stat = "identity")+
        ylab("Frequency Count")+
        xlab("Event Type")+
        theme(axis.text.x = element_text(angle=90, hjust=1))+
        ggtitle("Top 10 US Killers")+ theme(plot.title = element_text(hjust = 0.5))
```

### 3.2: Events that have the Greatest Economic Consequences

Melting data.table so that it is easier to put in bar graph format

``` {r}
econ_consequences <- melt(totalCostDT, id.vars="EVTYPE", variable.name = "Damage_Type")
head(econ_consequences, 5)

#create graph
ggplot(econ_consequences, aes(fill = Damage_Type, x = reorder(EVTYPE,-value), y = value))+
        geom_bar(position = "dodge", stat = "identity")+
        ylab("Cost (dollars)")+
        xlab("Event Type")+
        theme(axis.text.x = element_text(angle=90, hjust=1))+
        ggtitle("Top 10 US Storm Events causing Economic Consequences") + theme(plot.title = element_text(hjust = 0.5))

```



